DRAFT: Harmonized and trait aggregation paper

### 2020-11-20 Paper/Draft_harmonization_paper.tex
Update 20.11: Rough draft for discussion

The original RDA in Szöcs et al. \cite{szocs_effects_2014} indicated that downstream sites (high salinity) were characterised by the traits shredder, ovoviviparity, multivoltinism, long life cycle ($> 1$ year), and gill respiration and upstream sites (low salinity) were characterized by univoltinism, oviposition in clutches and short life cycle duration ($< 1$ year).

Using harmonized grouping features resulted in fewer traits that distinguish upstream and downstream sites in comparison to the original analysis (Figure \ref{fig:rda_species_scores} and Figure \ref{fig:boxplots_scores_on_constrained_axis}). According to the RDA of the trait composition, downstream sites were characterised by taxa with the traits multivoltinism and ovoviviparity and upstream sites were characterised by univoltine taxa that lay their eggs in an aquatic environment (aquatic eggs). The traits shredder, gill respiration and long life cycle duration did not characterise sites with high salinisation. Also, the trait short life cycle duration did not characterise upstream sites with low salinity.

Using at family-level aggregated traits from the harmonized dataset showed results similar to the original analysis (Figure \ref{fig:rda_species_scores} and Figure \ref{fig:boxplots_scores_on_constrained_axis}). The \textit{direct\_agg\textsubscript{mean}}, \textit{direct\_agg\textsubscript{median}}, and \textit{weighted\_agg} 
characterised the downstream sites with the same traits as the original analysis except that downstream sites were not characterised by the trait shredder. Upstream sites were characterised by the traits univoltinism, aquatic eggs and short life cycle duration. The trait aquatic eggs of the harmonized grouping feature oviposition has been derived by amalgamating the trait oviposition in clutches and other related traits (Table \ref{tab:traits_harmonization}). The same results were obtained when re-analysing the data with traits aggregated by the \textit{stepwise\_agg\textsubscript{mean}} and \textit{stepwise\_agg\textsubscript{median}}, with the exception that none of the life cycle traits characterised upstream or downstream sites. Thus, in our comparison the direct and weighted family-level aggregation methods yielded to the least change in species scores of all methods compared to the original results.

@@

% Summarize briefly results and synthesize (with Simulation)  
Our results show that aggregation of traits by the median compared to aggregating by the mean is often closer to the assigned traits at family level, i.e. it yields less differing cases. 
% This part needs to be improved:
% However, aggregation methods using the median tend to produce greater differences between aggregated and assigned traits.
The different weighting methods, i.e. aggregating stepwise, direct or weighted, did not seem to have a strong impact, because the distribution of absolute trait affinity differences to assigned traits obtained by the \textit{direct\_agg\textsubscript{mean}}, \textit{stepwise\_agg\textsubscript{mean}} and \textit{weighted\_agg} were relatively similar in both tested datasets.  %as well as for direct\agg (median) and stepwise\_agg (median)
Instead, using the mean or the median changed the distribution of differences in trait affinities between aggregated and assigned traits.
% Wording!
A reason that using different weightings for the taxa when aggregating doesn't seem to have a big influence is on the one hand, that the used taxa had only a low number of genera in the used trait datasets or don't have species entries at all. From the taxa that were compared from the North American trait dataset to the assigned traits at family-level, 13.8 \%, 61.7 \% were resolved on genus-level. 52.1 \% had 5 or less genera and 12.8 \% contained just one genus (Figure S\ref{fig:tax_hierarchy_NOA}). In the Australian dataset 21.4 \% of the compared taxa had trait information resolved on family-level, 39.5 \% resolved on genus level, and 39.1 \% resolved on species-level. 67.73 \% of the taxa contain 5 or less genera and 39.6 \% just contained one genus (Figure S\ref{fig:tax_hierarchy_AUS}). The simulation results indicated that the choice of the aggregation method for small numbers of genera per family (5) can be important when one genus has a much larger number of species compared to other genera within a particular family. 
% How often in the trait datasets?
However, inspecting the simulated datasets revealed that often aggregated traits only differ slightly obtained from the different aggregation methods .

@@

% Of course all the comparisons have no "generellen Gültigkeisanspruch" - what's the true aggregated trait affinity characterising a particular family? Expert assignments maybe also biased based on experience with a particular group of taxa within a family.  
Evaluation of the differences between aggregated and assigned traits is difficult because it remains unclear what the true value of a particular trait for a particular organism on family-level is. It is indeed questionable whether the aggregation of trait information on species or genus-level to a point estimate on family-level can be an adequate description of a particular trait at all, especially for traits with a high variability. In fact, traits can vary strongly on more precise taxonomic levels than the family-level. For example, Monaghan and Soares found a high dispersion of trait affinities at genus-level within families in the Tachet database for traits of the grouping features body size, flow preference, and reproduction \cite{monaghan_improving_2013}. Furthermore, studies focusing on the lability of traits, i.e. how much traits are unconstrained by phylogeny, found that traits of ecological preferences (e.g. thermal preference), body size, resistance forms, and to a lesser extent feeding mode are labile, and thus possibly highly variable \cite{poff_functional_2006,wilkes_traitbased_2020}. Thus, it is likely that assigned traits and aggregated traits at family-level will especially differ in affinities for the aforementioned traits. Our analysis could confirm this partially. The highest possible difference of 1 between aggregated trait affinities and assigned trait affinities (in total in 6.9 \% of all cases) for the Australian dataset occurred most often for body size large and medium (42.5 \% and 39.2 \% of all cases with a difference of 1, respectively) across all comparisons. Both traits had also the highest mean and standard deviation in the absolute difference of trait affinities across all tested comparisons (mean size medium 0.5, sd 0.29; mean size large 0.59, sd 0.32). For the taxa compared from the North American dataset, the highest possible difference in trait affinities occurred most often for the trait body size medium (15.8 \% of all cases with a maximum difference), followed by the traits respiration via plastron and spiracle and respiration via tegument (both 13 \% of all cases with a maximum difference). Respiration via plastron and spiracle showed the highest mean and standard deviation in the absolute difference of trait affinities across all comparisons (mean 0.71, sd 0.33), followed by respiration via tegument (mean 0.5, sd 0.34) and body size medium (mean 0.49, sd 0.3).

@@

% Further ideas: 
%   - Differences across grouping features
%   - Still some similarities between compared datasets -> Where does the classification come from?
%     Few sources used in NOA also used in AUS:
%     AUS: Merrit & Cummins 1978/1984 for feeding mode for a few genera and families 
%     NOA: Merritt, Richard W., Cummins, Berg 2008 -> but not clear if for feeding mode as well; according to trait   %     description "Cummins, K. W. Trophic Relations of Aquatic Insects" was used
%   - Could not include all data due to data gaps -> much needed to be filled -> would maybe change the picture?

@@

% Were the traits promoted by the harmonized dataset on the most extreme points on the conductivity axis?
% Large scale studies?

>>>>

The original RDA in Szöcs et al. \cite{szocs_effects_2014} indicated that downstream sites (high salinity) were characterised by the traits shredder, ovoviviparity, multivoltinism, long life cycle duration ($> 1$ year), and gill respiration and upstream sites (low salinity) were characterized by univoltinism, oviposition in clutches and short life cycle duration ($< 1$ year).

Using harmonized grouping features resulted in less traits that distinguish upstream and downstream sites in comparison to the original analysis (Figure \ref{fig:rda_species_scores} and Figure \ref{fig:boxplots_scores_on_constrained_axis}). According to the RDA of the trait composition, downstream sites were characterised by taxa with the traits multivoltinism, ovoviviparity and long life cycle duration. Upstream sites were characterised by univoltine taxa with a short life cycle duration that lay their eggs in an aquatic environment (aquatic eggs). The trait aquatic eggs of the harmonized grouping feature oviposition has been derived by amalgamating the trait oviposition in clutches and other related traits (Table \ref{tab:traits_harmonization}). The traits shredder and gill respiration did not characterise sites with high salinisation.

Using at family-level aggregated traits from the harmonized dataset showed also results similar to the original analysis but with fewer traits distinguishing upstream and downstream sites (Figure \ref{fig:rda_species_scores} and Figure \ref{fig:boxplots_scores_on_constrained_axis}). The \textit{direct\_agg\textsubscript{mean}}, \textit{direct\_agg\textsubscript{median}}, and \textit{weighted\_agg} 
characterised the downstream sites with the same traits as the original analysis except that downstream sites were not characterised by the trait shredder. Upstream sites were characterised by the traits univoltinism, aquatic eggs and short life cycle duration. The same results were obtained when re-analysing the data with traits aggregated by the \textit{stepwise\_agg\textsubscript{mean}} and \textit{stepwise\_agg\textsubscript{median}}, with the exception that none of the life cycle traits characterised upstream or downstream sites. Thus, in our comparison aggregating by \textit{direct\_agg\textsubscript{mean}}, \textit{direct\_agg\textsubscript{median}}, and \textit{weighted\_agg} yielded to the least change in interpretation of the RDA results compared to the original results.

@@

% Summarize briefly results and synthesize (with Simulation)  
Our results show that aggregation of traits by the median compared to aggregating by the mean is often closer to the assigned traits at family level, i.e. it yields less differing cases. However, in those cases where differences in the trait aggregation methods that use the median occurred, these differences were greater compared to the aggregation methods that used the mean. The different weighting methods, i.e. aggregating stepwise, direct or weighted, did not seem to have a strong impact, because the distribution of absolute trait affinity differences to assigned traits obtained by the \textit{direct\_agg\textsubscript{mean}}, \textit{stepwise\_agg\textsubscript{mean}} and \textit{weighted\_agg} were relatively similar in both tested datasets. This also applies to the distributions of absolute trait affinity differences obtained by \textit{direct\_agg\textsubscript{median}} and \textit{stepwise\_agg\textsubscript{median}}. 
Instead, using the mean or the median changed the distribution of differences in trait affinities between aggregated and assigned traits. A possible reason why the use of different weightings for taxa appears to have little impact on trait aggregation is that a considerable portion of taxa used have a low numbers of genera or species entries. From the taxa that were compared from the North American trait dataset to the assigned traits at family-level, 13.8 \% were resolved on family-level and 61.7 \% were resolved on genus-level. 52.1 \% had 5 or less genera and 12.8 \% contained just one genus (Figure S\ref{fig:tax_hierarchy_NOA}). In the Australian dataset 21.4 \% of the compared taxa had trait information resolved on family-level, 39.5 \% resolved on genus level, and 39.1 \% resolved on species-level. 67.73 \% of the taxa contain 5 or less genera and 39.6 \% just contained one genus (Figure S\ref{fig:tax_hierarchy_AUS}). The simulation results indicated that the choice of the aggregation method for small numbers of genera per family (5) can be important when one genus has a much larger number of species compared to other genera within a particular family. % How often in the trait datasets?
In general however, inspecting the simulated datasets revealed that often aggregated traits obtained from the different aggregation methods only differ slightly.

@@

% Of course all the comparisons have no "generellen Gültigkeisanspruch" - what's the true aggregated trait affinity characterising a particular family? Expert assignments maybe also biased based on experience with a particular group of taxa within a family.  
Evaluation of the differences between aggregated and assigned traits is difficult because it remains unclear what the true value of a particular trait for a particular family is. It is indeed questionable whether the aggregation of trait information on species or genus-level to a point estimate on family-level can be an adequate description of a particular trait at all, especially for traits with a high variability. In fact, traits can vary strongly on more precise taxonomic levels than the family-level. For example, Monaghan and Soares found a high dispersion of trait affinities for genera within families in the Tachet database for traits of the grouping features body size, flow preference, and reproduction \cite{monaghan_improving_2013}. Furthermore, studies focusing on the lability of traits, i.e. how much traits are unconstrained by phylogeny, found that traits of ecological preferences (e.g. thermal preference), body size, resistance forms, and to a lesser extent feeding mode are labile, and thus possibly highly variable \cite{poff_functional_2006,wilkes_traitbased_2020}. Thus, it is likely that assigned traits and aggregated traits at family-level will especially differ in affinities for the aforementioned traits. Our analysis could confirm this partially. The highest possible difference of 1 between aggregated trait affinities and assigned trait affinities (in total in 6.9 \% of all cases) for the Australian dataset occurred most often for body size large and medium (42.5 \% and 39.2 \% of all cases with a difference of 1, respectively) across all comparisons. Both traits had also the highest mean and standard deviation in the absolute difference of trait affinities across all tested comparisons (size medium: mean 0.5, sd 0.29; size large: mean 0.59, sd 0.32). For the taxa compared from the North American dataset, the highest possible difference in trait affinities occurred most often for the trait body size medium (15.8 \% of all cases with the maximum difference of 1), followed by the traits respiration via plastron and spiracle and respiration via tegument (both 13 \% of all cases with a maximum difference of 1). Respiration via plastron and spiracle showed the highest mean and standard deviation in the absolute difference of trait affinities across all comparisons (mean 0.71, sd 0.33), followed by respiration via tegument (mean 0.5, sd 0.34) and body size medium (mean 0.49, sd 0.3). 
% What to do with these traits? Also briefly talk about traits that are more consistent?

@@

% Further ideas: 
%   - Differences across grouping features
%   - Still some similarities between compared datasets -> Where does the classification come from?
%     Few sources used in NOA also used in AUS:
%     AUS: Merrit & Cummins 1978/1984 for feeding mode for a few genera and families 
%     NOA: Merritt, Richard W., Cummins, Berg 2008 -> but not clear if for feeding mode as well; according to trait description "Cummins, K. W. Trophic Relations of Aquatic Insects" was used for feeding mode
%   - Could not include all data due to data gaps -> much needed to be filled -> would maybe change the picture?

@@

The re-analysis showed that neither using harmonised grouping features nor aggregated traits changed drastically the results of the original study. The traits that did not respond to the salinity gradient (i.e. not on the extremes of the conductivity axis) in the re-analysis were those closest to the criterion that defined when a trait was associated with either high or low salinity (mahalanobis distance greater than the 97.5 \% quantile of the Chi-square distribution, see methods), namely feeding mode shredder, respiration gills, and life cycle duration traits. During harmonisation the trait shredder has been amalgamated based on three traits (Miner, xylophagus, and shredder). As a result, the trait affinities in the original dataset have a higher mean and standard deviation than in the aggregated and harmonized datasets (Table S\ref{tab:SI_resp_traits_summary_stats}).
% TODO: Check naming of tables! 
Thus, the signal in the original data for the trait shredder has presumably been weakened by the harmonization, which might be the reason that shredders showed a less strong reaction to the salinity gradient.

% Resp gil complicated: Original data has a lot of zeros, not in the harmonized & aggregated datasets
% Aggregated data look similar to the harmonized ones, but resp gil is here one of the "extreme" traits 
% on the conductivity axis

% Why did some aggregation methods promote life cycle duration and others not?
% Mean, median and sd's are all relatively similar across tested datasets for these traits

% Large scale studies?

### 2020-11-20 Paper/Results/SI.tex
Update 20.11: Rough draft for discussion

>>>>

\subsection*{Re-analysis of Szöcs et al. using harmonized and aggregated grouping features}

\begin{table}[ht]
    \centering
    \caption{Mean, median and standard deviation of the traits that were responsive to the salinity gradient in the original study but not when re-analysis with the harmonized dataset trait dataset.} 
    \label{tab:SI_resp_traits_summary_stats}
    \begin{tabular}{llccc}
    \toprule[.1em]
    Dataset & Trait & Mean & Median & SD \\ 
    \toprule[.1em]
    Stepw\_median & Shredder & 0.20 & 0.14 & 0.25 \\ 
      Stepw\_mean & Shredder & 0.18 & 0.12 & 0.22 \\ 
      Direct\_median & Shredder & 0.21 & 0.14 & 0.25 \\ 
      Direct\_mean & Shredder & 0.19 & 0.14 & 0.22 \\ 
      Weighted & Shredder & 0.19 & 0.14 & 0.22 \\ 
      Harmonized; not\_aggregated & Shredder & 0.18 & 0.12 & 0.24 \\ 
      Original & Shredder & 0.25 & 0.14 & 0.32 \\ 
      Stepw\_median & Gills & 0.30 & 0.27 & 0.32 \\ 
      Stepw\_mean & Gills & 0.29 & 0.22 & 0.32 \\ 
      Direct\_median & Gills & 0.30 & 0.30 & 0.32 \\ 
      Direct\_mean & Gills & 0.30 & 0.30 & 0.32 \\ 
      Weighted & Gills & 0.30 & 0.30 & 0.32 \\ 
      Harmonized; not\_aggregated & Gills & 0.30 & 0.25 & 0.32 \\ 
      Original & Gills & 0.28 & 0.00 & 0.33 \\ 
      Stepw\_median & Short life cycle & 0.64 & 0.75 & 0.39 \\ 
      Stepw\_mean & Short life cycle & 0.64 & 0.79 & 0.39 \\ 
      Direct\_median & Short life cycle & 0.67 & 0.75 & 0.37 \\ 
      Direct\_mean & Short life cycle & 0.67 & 0.79 & 0.38 \\ 
      Weighted & Short life cycle & 0.67 & 0.79 & 0.38 \\ 
      Harmonized; not\_aggregated & Short life cycle & 0.64 & 0.75 & 0.40 \\ 
      Original & Short life cycle & 0.64 & 0.75 & 0.40 \\ 
      Stepw\_median & Long life cylce & 0.36 & 0.25 & 0.39 \\ 
      Stepw\_mean & Long life cylce & 0.36 & 0.21 & 0.39 \\ 
      Direct\_median & Long life cylce & 0.33 & 0.25 & 0.37 \\ 
      Direct\_mean & Long life cylce & 0.33 & 0.21 & 0.38 \\ 
      Weighted & Long life cylce & 0.33 & 0.21 & 0.38 \\ 
      Harmonized; not\_aggregated & Long life cylce & 0.36 & 0.25 & 0.40 \\ 
      Original & Long life cylce & 0.36 & 0.25 & 0.40 \\ 
    \bottomrule
    \end{tabular}
\end{table}
